---
title: "DBA5104 igraph Tutorial in R"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This markdown is a short tutorial on how to use igraph in R. When using igraph,
R is preferred over python as R is more stable and has more comprehensive 
functions.

You can check out the documentation: https://igraph.org/r/doc/

## Initialization
```{r}
#install.packages("igraph")

library(igraph)
library(igraphdata)
library(sand)
```

### Creating Network Graphs
```{r}
# For small, toy graphs, the function graph_from_literal can be used, 
# specifying the edges in a symbolically literal manner

g <- graph_from_literal(1-2, 1-3, 2-3, 2-4, 3-5, 4-5, 4-6, 4-7, 5-6, 6-7)
```

Number of vertices
```{r}
V(g)
```

Number of edges
```{r}
E(g)
```

```{r}
#  same information, combined and in a slightly different format, is recovered
print_all(g)
```

Visual representation
```{r}
plot(g)
```

Adjacency matrix
```{r}
 as_adjacency_matrix(g)
```

### Graph Concepts
Check for weighted graph
```{r}
is_weighted(g)
```

Check for simple graph (i.e. no self-loop or multi-edges)
```{r}
is_simple(g)
```

Check for neighbour of a certain node
```{r}
# Neighbours of node 5 in the graph g

neighbors(g,5)
```

Check for the degrees of the nodes in the graph
```{r}
degree(g)
```

Check if the graph is connected
```{r}
is_connected(g)
```

Check for clusters in graph (if the graph is connected - 1 component)
```{r}
clusters(g)
```

Check for diameter (Longest shortest path)
```{r}
diameter(g, weights=NA)
```

Directed graph in graph with pairs
```{r}
g2 <- graph( c("A", "B", "B", "C", "B", "C", "A", "A"),
isolates=c("D") )
plot(g2)
```

Directed edges in graph_from_literal are indicated using a minus/plus
convention
```{r}
dg <- graph_from_literal(1-+2, 1-+3, 2++3)
plot(dg)
```

### Descriptive Analysis of Network Graph Characteristics
Initialize toy dataset
```{r}
head(elist.lazega)
head(v.attr.lazega)
g.lazega=graph_from_data_frame(elist.lazega,directed="FALSE",vertices = v.attr.lazega)
```

Betweenness centrality of every node in the graph
```{r}
betweenness(g.lazega)
```

Eigenvector centrality of every node in the graph
```{r}
eigen_centrality(g.lazega)$vector
```

Transitivity (i.e. clustering coefficient) of the graph
```{r}
transitivity(g.lazega)
```

### Star Wars dataset
Obtained from 
https://github.com/pablobarbera/data-science-workshop/blob/master/sna/data/

Description of dataset
```{r}
edges <- read.csv("C:/Users/Eugene/Downloads/star-wars-network-edges.csv")
nodes <- read.csv("C:/Users/Eugene/Downloads/star-wars-network-nodes.csv")
head(edges)
```
```{r}
head(nodes)
```

Create an igraph object from the above raw data.
```{r}
g <- graph_from_data_frame(d=edges, vertices=nodes, directed=FALSE) 
#d describes the edges of the network
g
```

```{r}
V(g)
```

```{r}
E(g)
```

Character names
```{r}
V(g)$name
```

Edge weights
```{r}
E(g)$weight
```

```{r}
par(mar=c(0,0,0,0))
plot(g)
```

Visualise it better? To see all the available plotting options, you can check ?igraph.plotting.
```{r}
par(mar=c(0,0,0,0))
plot(g,
vertex.color = "grey", # change color of nodes
vertex.label.color = "black", # change color of labels
vertex.label.cex = .75, # change size of labels to 75% of original size
edge.curved=.25, # add a 25% curve to the edges
edge.color="grey20") # change edge color to grey
```

Change based on which side they belong to:
```{r}
# create vectors with characters in each side
dark_side <- c("DARTH VADER", "MOTTI", "TARKIN")
light_side <- c("R2-D2", "CHEWBACCA", "C-3PO", "LUKE", "CAMIE", "BIGGS",
"LEIA", "BERU", "OWEN", "OBI-WAN", "HAN", "DODONNA",
"GOLD LEADER", "WEDGE", "RED LEADER", "RED TEN", "GOLD FIVE")
other <- c("GREEDO", "JABBA")
# node we'll create a new color variable as a node property
V(g)$color <- NA
V(g)$color[V(g)$name %in% dark_side] <- "red"
V(g)$color[V(g)$name %in% light_side] <- "gold"
V(g)$color[V(g)$name %in% other] <- "grey20"
vertex_attr(g)
```

```{r}
par(mar=c(0,0,0,0)); plot(g)
legend(x=.75, y=.75, legend=c("Dark side", "Light side", "Other"),
pch=21, pt.bg=c("red", "gold", "grey20"), pt.cex=2, bty="n")
```

Up to now, each time we run the plot function, the nodes appear to be in a 
different location. Why? Because it’s running a probabilistic function trying to
locate them in the optimal way possible. However, we can also specify the layout
for the plot; that is, the (x,y) coordinates where each node will be placed. 
igraph has a few different layouts built-in, that will use different algorithms 
to find an optimal distribution of nodes. The following code illustrates some 
of these:
```{r}
par(mfrow=c(2, 3), mar=c(0,0,1,0))
plot(g, layout=layout_randomly, main="Random")
plot(g, layout=layout_in_circle, main="Circle")
plot(g, layout=layout_as_star, main="Star")
plot(g, layout=layout_as_tree, main="Tree")
plot(g, layout=layout_on_grid, main="Grid")
plot(g, layout=layout_with_fr, main="Force-directed")
```

## References
Statistical Analysis of Network Data with R - Eric D. Kolaczyk,
Gábor Csárdi

### More help:
Use command: ?igraph
or visit https://cran.r-project.org/web/packages/igraph/igraph.pdf